---
title: "Data Analysis Assignment #2 (75 points total)"
author: "Woods, Stephen"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv", sep = ",", stringsAsFactors = TRUE)
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

***#### Section 1: (5 points) ####***

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a, fig.width = 12}

par(mfrow = c(1,2))
hist(mydata$RATIO, col = "indianred", main = "Histogram of Ratio", xlab = "RATIO", ylab = "count", breaks = "FD")
qqnorm(mydata$RATIO, col = "indianred", main = "QQ Plot of Ratio", xlab = "theoretical", ylab = "sample")
qqline(mydata$RATIO)
par(mfrow = c(1,1))

skewness(mydata$RATIO)
kurtosis(mydata$RATIO)

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b, fig.width = 12, fig.height = 8}

mydata$L_RATIO <- log10(mydata$RATIO)

layout(matrix(c(1,2,3,3), 2, 2, byrow = TRUE))
hist(mydata$L_RATIO, col = "indianred", main = "Histogram of L_Ratio", xlab = "L_RATIO", ylab = "count", breaks = "FD")
qqnorm(mydata$L_RATIO, col = "indianred", main = "QQ Plot of L_Ratio", xlab = "theoretical", ylab = "sample")
qqline(mydata$L_RATIO)


skewness(mydata$L_RATIO)
kurtosis(mydata$L_RATIO)

boxplot(mydata$L_RATIO ~ mydata$CLASS, col = "cadetblue", main = "L_RATIO by CLASS", xlab = "CLASS", ylab = "L_RATIO")

```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
bartlett.test(mydata$L_RATIO ~ mydata$CLASS)
bartlett.test(mydata$RATIO ~ mydata$CLASS)

qchisq(.95, 4)
```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: (L_RATIO exhibits better conformance to a normal distribution with skewness and kurtosis closer to that of a normal distribution. The histogram of ratio shows right skew, while the histogram of L_Ratio shows no visual skewness. Bartlett's test of homogeneity of variance using the L_Ratio shows no significant variance across classes, with a P-value of .5267 and a k-squared value below the critical value. The Bartlett test using the Ratio variable shows the assumption of homogeneity is false. )***


***#### Section 2 (10 points) ####***

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}

summary(aovmdint<- aov(L_RATIO ~ CLASS*SEX, data = mydata))
summary(aov_md1 <- aov(L_RATIO ~ CLASS + SEX, data = mydata))

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: (The interaction between CLASS and SEX does not significantly affect the variance of L_Ratio, therefore we do not want to include that interaction term in future regression models. L_Ratio and SEX both independently explain significant variance among L_Ratio means. Further analysis is needed to determine where the significant difference lies between levels of CLASS and SEX. )***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(aov_md1)

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: (Between all CLASS levels other than A2-A1, there is a significant difference in L_Ratio. This indicates that L_Ratio increases as the class of an abalone increases. The larger the gap in CLASS levels, the lower the P-Value. Between all SEX levels other than M-F, there is a significant difference in L_Ratio. With a p-adj value of .94, we can confidently conclude the difference between Male and Female categories have no significant affect on L_Ratio. Therefore, we can combine Male and Female categories into one "Adult" category to simplify our data. )***


***####  Section 3: (10 points) ####***

(3)(a1) Here, we will combine "M" and "F" into a new level, "ADULT". The code for doing this is given to you. For (3)(a1), all you need to do is execute the code as given.

```{r Part_3a1}
# Here, we show how to define the new variable TYPE using only base R functions:

mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
table(mydata$TYPE)


```

(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2, fig.width = 12}

par(mfrow = c(1,2))

hist(mydata$VOLUME[mydata$TYPE == "I"], col = "indianred", main = "Histogram of Infant TYPE", xlab = "VOLUME", ylab = "count", breaks="FD")
hist(mydata$VOLUME[mydata$TYPE == "ADULT"], col = "cadetblue", main = "Histogram of Adult TYPE", xlab = "VOLUME", ylab = "count", breaks="FD")
mtext("A1", side = 1, outer = TRUE, line = 2.2)


par(mfrow = c(1,1))

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: (The infant distribution shows a strong right skew with most Infants having volumes between 0-400. Though, a couple infants have volumes greater than 700. The distribution of Adult VOlumes is more even. With adult volumes commonly falling in the 0-400 range as well, there is going to be some difficulty separating infants from adults based on volume.)***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b, fig.width = 12, fig.height = 8}

library(ggthemes)
library(ggpubr)
library(ggplot2)

mydata$L_SHUCK <- log10(mydata$SHUCK)
mydata$L_VOLUME <- log10(mydata$VOLUME)


  scat1 <- ggplot(data = mydata)+ aes(x = VOLUME, y = SHUCK) + geom_point(aes(color = CLASS)) + theme(legend.position = "bottom")
  scat2 <- ggplot(data = mydata)+ aes(x = L_VOLUME, y = L_SHUCK) + geom_point(aes(color = CLASS)) + theme(legend.position = "bottom")
 
  scat3 <- ggplot(data = mydata)+ aes(x = VOLUME, y = SHUCK) + geom_point(aes(color = TYPE)) + theme(legend.position = "bottom")
  scat4 <- ggplot(data = mydata)+ aes(x = L_VOLUME, y = L_SHUCK) + geom_point(aes(color = TYPE)) + theme(legend.position = "bottom")
 
figure <- ggarrange(scat1, scat2, scat3, scat4,
          font.label = list(size = 10),
          #vjust = .3,
          #hjust = -1.2,
          ncol = 2, nrow = 2,
          align = "v")

annotate_figure(figure,
               top = text_grob("VOLUME + SHUCK by CLASS and TYPE", size = 14))

```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: (Log10-transformation appears to reduce the variability of both plots differentiated by CLASS and TYPE and makes differences between CLASSES and TYPES more visible. It also shifts the bulk of the data points from the lower left quadrant to the upper right quadrant. Lower CLASSES, especially CLASS A1, are more dominant in the lower left quadrant of the plots, with classes increasing as we move to the upper right quadrant of the plots. CLASS differences are less apparent between between A3 and A5. A similar trend is present for TYPE - infants dominate the lower left quadrant and adults dominate the upper rate quadrant. With less variance than SHUCK and VOLUME, a linear regression model using L_SHUCK and L_VOLUME as explanatory variables should result in lower residual values.  )***


***####   Section 4: (5 points) ####***

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. You are given code in (4)(a1) to reclassify the infants in classes A4 and A5 as ADULTS. 

```{r Part_4a1}

mydata$TYPE[mydata$CLASS == "A4" | mydata$CLASS == "A5"] <- "ADULT"
table(mydata$TYPE)

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}


mydata$TYPE <- relevel(x = mydata$TYPE, ref = "I")

#levels(mydata$TYPE)

summary(lr_md1 <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata))

```

**Essay Question:  Interpret the trend in CLASS levelcoefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: (As class level increases, the coefficient estimates decrease slightly. This suggests that L_SHUCK decreases as abalone class increases, not show in the charts in 3b. From the charts presented in 3.b, we know that class A1 has the most distinct L_SHUCK, so it does make sense that regression coefficients for further classes are not very large or materially significant. The boxplots in 1.b do show that :_Ratio decreases as class increases, with no visible difference from class A1 to A2. This matches the model summary, where the coefficient estimate from A1 to A2 is not significant.  )***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: (With a coefficient estimate of .021, type does not appear to be a materially significant predictor of L_SHUCK. L_Volume and most class types have larger coefficients than Type. We will lose one degree of freedom by including type as a predictor, so we can determine whether it adds value in future analysis.)***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

***#### Section 5: (5 points) ####***

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5am, fig.width = 12}


par(mfrow = c(1,2))
hist(lr_md1$residuals, col = "indianred", main = "Histogram of Model Residuals", xlab = "Model Residuals", ylab = "count", breaks = "FD")
qqnorm(lr_md1$residuals, col = "indianred", main = "QQ Plot of Model Residuals", xlab = "theoretical", ylab = "sample")
qqline(lr_md1$residuals)
par(mfrow = c(1,1))

skewness(lr_md1$residuals)
kurtosis(lr_md1$residuals)

```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b, fig.width = 12, fig.height = 8}
  scat5.1 <- ggplot(data = mydata)+ aes(x = L_VOLUME, y = lr_md1$residuals) + geom_point(aes(color = CLASS)) + theme(legend.position = "bottom")
  scat5.2 <- ggplot(data = mydata)+ aes(x = L_VOLUME, y = lr_md1$residuals) + geom_point(aes(color = TYPE)) + theme(legend.position = "bottom")
 
  box5.1 <- ggplot(data = mydata)+ aes(x = CLASS, y = lr_md1$residuals) + stat_boxplot(color = "black", fill = "indianred")
  box5.2 <- ggplot(data = mydata)+ aes(x = TYPE, y = lr_md1$residuals) + stat_boxplot(color = "black", fill = "cadetblue")
 
figure <- ggarrange(scat5.1, scat5.2, box5.1, box5.2,
          font.label = list(size = 10),
          #vjust = .3,
          #hjust = -1.2,
          ncol = 2, nrow = 2,
          align = "v")

annotate_figure(figure,
               top = text_grob("Residuals by VOLUME + CLASS and TYPE", size = 14))

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: (The histogram of residuals reflects a normal distribution with near zero skewness and little kurtosis. Similarly, the QQ plot does not show many points significantly far from a normal line. The residual scatterplots both show clustering, but there is no other clear trent indicating residuals are not evenly dispersed. Similarly, the boxplots by both class and type have means near zero and all look fairly indistinguishable from one another. All of these charts indicate homogeneity of variance, we do not reject the null hypothesis. This strongly indicates L_VOLUME or VOLUME may be useful for making harvesting decisions.)***

-----

Harvest Strategy:

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.The Management needs to make a decision to implement 1 rule that meets the business goal.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

***#### Section 6: (5 points) ####***

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

head(volume.value)
head(prop.infants)
head(prop.adults)



```

(6)(b) Our first "rule" will be protection of all infants. We want to find a volume cutoff that protects all infants, but gives us the largest possible harvest of adults. We can achieve this by using the volume of the largest infant as our cutoff. You are given code below to identify the largest infant VOLUME and to return the proportion of adults harvested by using this cutoff. You will need to modify this latter code to return the proportion of infants harvested using this cutoff. Remember that we will harvest any individual with VOLUME greater than our cutoff.

```{r Part_6b}
# Largest infant volume
(max_inf_vol <- max(mydata$VOLUME[mydata$TYPE == "I"]))  # [1] 526.6383

# Proportion of adults harvested
(propadl <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) /
  total.adults)  # [1] 0.2476573


# Add code to calculate the proportion of infants harvested
(propinf <- sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) /
  total.infants)

# If we use the largest infant volume, we harvest approximately 24.8% of adults and 0%,
# as expected, of infants.

```

(6)(c) Our next approaches will look at what happens when we use the median infant and adult harvest VOLUMEs. Using the median VOLUMEs as our cutoffs will give us (roughly) 50% harvests. We need to identify the median volumes and calculate the resulting infant and adult harvest proportions for both.

```{r Part_6c}
# Add code to determine the median infant volume:

(med_inf_vol <- median(mydata$VOLUME[mydata$TYPE == "I"]))

# Add code to calculate the proportion of infants harvested

(propinf1 <- sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol) /
  total.infants)

# Add code to calculate the proportion of adults harvested
(propadl1 <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol) /
  total.adults)

# If we use the median infant volume as our cutoff, we harvest almost 50% of our infants
# and a little more than 93% of our adults.

# Add code to determine the median adult volume:
(med_adl_vol <- median(mydata$VOLUME[mydata$TYPE == "ADULT"]))

# Add code to calculate the proportion of infants harvested
(propinf2 <- sum(mydata$VOLUME[mydata$TYPE == "I"] > med_adl_vol) /
  total.infants)

# Add code to calculate the proportion of adults harvested
(propadl2 <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_adl_vol) /
  total.adults)

# If we use the median adult volume as our cutoff, we harvest almost 50% of adults
# and approximately 2.4% of infants.

```

(6)(d) Next, we will create a plot showing the infant conserved proportions (i.e. "not harvested," the prop.infants vector) and the adult conserved proportions (i.e. prop.adults) as functions of volume.value. We will add vertical A-B lines and text annotations for the three (3) "rules" considered, thus far:  "protect all infants," "median infant" and "median adult." Your plot will have two (2) curves - one (1) representing infant and one (1) representing adult proportions as functions of volume.value - and three (3) A-B lines representing the cutoffs determined in (6)(b) and (6)(c).

```{r Part_6d, fig.width = 12, fig.height = 6}

plot_6d <- ggplot()+ aes() + geom_line(aes(volume.value, prop.infants, color = "indianred"), size = 1.5) +
  geom_line(aes(volume.value, prop.adults, color = "cadetblue"), size = 1.5) + 
  geom_vline(xintercept = max_inf_vol) +
  geom_vline(xintercept = med_inf_vol) +
  geom_vline(xintercept = med_adl_vol) +
  geom_hline(yintercept = .5, linetype = "dashed") +
  annotate("text", x = med_inf_vol+70, y = .45, label = "Med inf: 133.8") +
  annotate("text", x = med_adl_vol+70, y = .45, label = "Med adult: 384.6") +
  annotate("text", x = max_inf_vol+70, y = .95, label = "Largest inf: 526.6") +
  ggtitle("Proportion of Adults and Infants Protected") +
  xlab("Volume") + ylab("Proportion") +
  scale_color_manual(labels = c("Adults", "Infants"), values = c("blue", "red")) +
  labs(color = "") +
  theme(legend.position = c(.9, 0.7), legend.background = element_blank())

plot_6d
  

```

**Essay Question:  The two 50% "median" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: (Using the median infant volume as a cutoff point will result in harvesting 50% of infants and 93% of adults. Using median adult volume as a cutoff point will result in harvesting 2% of infants and 50% of adults. Using the median adult cutoff point will better preserve infants, promoting there will be more adults to harvest in the future.)***

-----


More harvest strategies:

This part will address the determination of a cutoff volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. In other words, we want to find the volume value such that the vertical distance between the infant curve and the adult curve is maximum. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

***#### Section 7: (10 points)  ####***

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}
difference <- ((1 - prop.adults) - (1 - prop.infants))
head(difference)

plot_7a <-ggplot()+ aes() + geom_line(aes(volume.value, difference, color = "indianred"), size = 1.5) +
  ggtitle("Difference vs Volume") +
  xlab("Volume") + ylab("Difference in Proportions Harvested") +
  theme(legend.position = "none")
 
plot_7a


```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

Include, side-by-side, the plot from (6)(d) but with a fourth vertical A-B line added. That line should intercept the x-axis at the "max difference" volume determined from the smoothed curve here.

```{r Part_7c, fig.width = 12, fig.height = 6}

max_diff <- volume.value[which.max(smooth.difference)]

plot_7a_2 <- plot_7a + geom_line(aes(volume.value, smooth.difference), linetype = "dashed", size = 1.5) +
  geom_vline(xintercept = max_diff)+
  annotate("text", label = paste('Volume =', round(max_diff, 3)), 
           x = max_diff + 20, y = 0.4, angle = 90)
  

plot_6d_2 <- ggplot()+ aes() + geom_line(aes(volume.value, prop.infants, color = "indianred"), size = 1.5) +
  geom_line(aes(volume.value, prop.adults, color = "cadetblue"), size = 1.5) + 
  geom_vline(xintercept = max_inf_vol) +
  geom_vline(xintercept = med_inf_vol) +
  geom_vline(xintercept = med_adl_vol) +
  geom_hline(yintercept = .5, linetype = "dashed") +
  annotate("text", x = med_inf_vol+10, y = .39, label = "Med inf: 133.8", angle = 90) +
  annotate("text", x = med_adl_vol+10, y = .35, label = "Med adult: 384.6", angle = 90) +
  annotate("text", x = max_inf_vol+10, y = .97, label = "Largest inf: 526.6") +
  ggtitle("Proportion of Adults and Infants Protected") +
  xlab("Volume") + ylab("Proportion") +
  scale_color_manual(labels = c("Adults", "Infants"), values = c("blue", "red")) +
  labs(color = "") +
  theme(legend.position = c(.9, 0.7), legend.background = element_blank())+
  
  
  geom_vline(xintercept = max_diff)+
  annotate("text", label = paste('Max diff:', round(max_diff, 1)), 
           x = max_diff + 10, y = 0.7, angle = 90)


figure <- ggarrange(plot_7a_2, plot_6d_2,
          font.label = list(size = 10),
          vjust = .3,
          hjust = -1.2,
          ncol = 2, nrow = 1,
          align = "v")

figure


```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions. We will actually calculate these proportions in two ways:  first, by 'indexing' and returning the appropriate element of the (1 - prop.adults) and (1 - prop.infants) vectors, and second, by simply counting the number of adults and infants with VOLUME greater than the vlume threshold of interest.

Code for calculating the adult harvest proportion using both approaches is provided.

```{r Part_7d}

#(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
# OR,
(propadl3 <- sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.adults) # [1] 0.7416332

(propinf3 <- (1 - prop.infants)[which.max(smooth.difference)])

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

***####  Section 8: (10 points)  ####***

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions. You may use either the 'indexing' or 'count' approach, or both.

```{r Part_8a}

(zero_a1_inf <- volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1]) # [1] 206.786

(propinf4 <- sum(mydata$VOLUME[mydata$TYPE == "I"] > zero_a1_inf) /
  total.infants)

(propadl4 <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > zero_a1_inf) /
  total.adults)



```

(8)(b) Next, append one (1) more vertical A-B line to our (6)(d) graph. This time, showing the "zero A1 infants" cutoff from (8)(a). This graph should now have five (5) A-B lines:  "protect all infants," "median infant," "median adult," "max difference" and "zero A1 infants."

```{r Part_8b, fig.width = 12}
plot_8b <- plot_6d_2 +
  geom_vline(xintercept = zero_a1_inf) +
  annotate("text", x = zero_a1_inf+10, y = .62, label = "Zero A1 inf: 206.8", angle = 90)

plot_8b

```


***#### Section 9: (5 points) ####***

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (6), (7) and (8) on this plot and label each. 

```{r Part_9, fig.width = 8.5}



points <- c(max_inf_vol, med_inf_vol, med_adl_vol, max_diff, zero_a1_inf)

adl_props <- c(propadl, propadl1, propadl2, propadl3, propadl4)
inf_props <- c(propinf, propinf1, propinf2, propinf3, propinf4)



ggplot(mapping = aes((1-prop.infants), (1-prop.adults))) + geom_line(color = "cadetblue", size = 1.5) +
  geom_abline(intercept = 0, slope = 1, color = 'darkred', linetype = 2) +
  geom_point(mapping = aes(inf_props, adl_props),
  shape = 21, size = 5) +
  annotate("text", label = paste('Protect all infants \nvol =', round(points[1], 1)), 
           x = 0.09, y = 0.25, size = 3) +
  annotate("text", label = paste('Median adult \nvol =', round(points[3], 1)), 
           x = 0.1, y = 0.5, size = 3) +
  annotate("text", label = paste('Max difference \nvol =', round(points[4], 1)), 
           x = 0.23, y = 0.7, size = 3) +
  annotate("text", label = paste('Zero A1 infants \nvol =', round(points[5], 1)), 
           x = 0.3, y = 0.9, size = 3) +
  annotate("text", label = paste('Median infant vol \nvol =', round(points[2], 1)), 
           x = 0.5, y = 0.85, size = 3) + 

  labs(title = "ROC Curve of Adult and Infant Harvest Proportions")
  




```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}

round(flux::auc(x = (1 - prop.infants), y = (1 - prop.adults)),3)
```


***#### Section 10: (10 points) ####***

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population

To calculate the total harvest proportions, you can use the 'count' approach, but ignoring TYPE; simply count the number of individuals (i.e. rows) with VOLUME greater than a given threshold and divide by the total number of individuals in our dataset.
 	
```{r Part_10} 	


tot_1 <- sum(mydata$VOLUME > max_inf_vol)/nrow(mydata)
tot_2 <- sum(mydata$VOLUME > med_inf_vol)/nrow(mydata)
tot_3 <- sum(mydata$VOLUME > med_adl_vol)/nrow(mydata)
tot_4 <- sum(mydata$VOLUME > max_diff)/nrow(mydata)
tot_5 <- sum(mydata$VOLUME > zero_a1_inf)/nrow(mydata)

total.cuts <- c(tot_1, tot_2, tot_3, tot_4, tot_5)



table10 <- data.frame(points, adl_props, inf_props, total.cuts)
rownames(table10) <- c("Protect all Infants", "Median Infants", "Median Adults", "Max Difference", "Zero A1 Infants")
colnames(table10) <- c("Volume", "TPR", "FPR", "totalHarvest")



round(table10, 2)




```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the five cutoffs determined in this assignment.**   

***Answer: (Our goal is to maximize the long term output of harvests by minimizing the harvesting of infants while maximizing the harvesting of adults. If we set the largest infant weight as the harvesting cutoff, no infants will be harvested, only 25% of adults will be harvested, and 18% of the total population will be harvested. With a high degree of overlap between infant and adult volumes, protecting all infants results in only harvesting 18% of the population. Using the median infant volume as a cutoff is less conservative, harvesting 81% of the population. But, harvesting 50% of our infant population will severely limit our future adult population. Using the median adult volume protects most infants but only harvests 37% of the total population. If we use the volume at which we maximize the difference between adults and infants harvested, we harvest 58% of the population while harvesting less than 20% of infants. Harvesting zero A1 infants yields a total harvest of 68% but harvests nearly 30% of all infants.)***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for  implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer: (When presenting analytic outcomes to management, I would provide an ultimate recommendation while outlining the the trade offs of various choices. My ultimate recommendation would be based on further analysis which would model the total volume of abalones harvested over a 10 year period. The cutoff point that yields the maximum total volume harvested over that 10 year period would be selected. This would require research and probably assumptions into how long it takes abalones to age. But, I would prefer to give management the ability to make a decision on more than a one year outlook. In presenting my analysis, I would highlight this is based on a sample of around 1,000 abalones. Abalones are hard to classify, and some outliers in the data point to the possibility of miss classification. Based on the current analysis, I would recommend the max difference cutoff point is used to promote a relatively large harvest while protecting most infants. For future studies of this type, I suggest collecting more features from each abalone to see if better predictors of abalone type exist. I would also reoccomend that samples are taken from different geographies to see if any geographical differences exist.)***